---
title: "Getting Started: Fitting sGP to a Synthetic Dataset"
author: "Ziang Zhang"
date: "2024-11-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE, warning=FALSE}
library(BayesGP)
library(tidyverse)
library(parallel)
source("code/functions.R")
```


## A Synthetic Dataset

In this tutorial, we introduce the basic steps to fit a sGP model using the seasonal B-spline approximation introduced in [Zhang et al, 2024](https://arxiv.org/abs/2305.09914).

For illustration, we use one of the synthetic datasets described in the main paper.
The dataset and its corresponding true function are shown in the plots below.

```{r}
n <- 500
location_of_interest <- seq(0, 10, length.out = 500)
true_f <- function(x){
  if(x < 2){
    return(2*sin(2 * 2 * pi * x) * (3-x))
  } else if (x > 2 && x < 4){
    return(2*sin(2 * 2 * pi * x))
  } else{
    return(2*sin(2 * 2 * pi * x) * (log(x-3) + 1))
  }
}
true_f <- Vectorize(true_f)

set.seed(123)
data <- simulate_data_poisson(func = true_f, n = n, sigma = 0.5, region = c(0,10), offset = 0)

par(mfrow = c(1,2))
plot(data$x, data$y, type = "p", col = "black", 
     pch = 20, cex = 0.5,
     ylab = "y", xlab = "x")
lines(location_of_interest, exp(true_f(location_of_interest)), col = "red", lwd = 2)
plot(location_of_interest, true_f(location_of_interest),
     type = "l", col = "black",
     pch = 20, cex = 0.5,
     ylab = "y", xlab = "x")
par(mfrow = c(1,1))
```


The hierarchical model we consider is as follows:

\begin{equation}
\begin{aligned}
Y_i|x_i,\xi_i &\sim \text{Poisson}(\lambda_i), \\
\text{log}(\lambda_i) &= \beta_0 + g(x_i) + \xi_i, \\
g(x) &\sim \text{sGP}_{\alpha}(\sigma_x), \\
\xi_i &\sim \text{N}(0, \sigma_\xi^2).
\end{aligned}
\end{equation}

We assume the sGP prior has a frequency of $2$ ($\alpha = 4\pi$), and use an Exponential prior for the one-step PSD $\sigma(1)$ defined as:
\[
\text{P}(\sigma(1) > 2) = 0.1.
\]

The standard deviation $\sigma_\xi$ of the observation-level random intercept that accounts for the overdispersion, follows an Exponential prior with a median of $1$.
All the fixed effects (including the boundary effects of the sGP) are assigned independent normal priors with zero mean and variance $1000$.

To make the computation more efficient, we will use $10$ equally spaced knots to define the B-spline basis, which will then be used to approximate the sGP prior.


## Inference using `BayesGP`

To make approximate Bayesian inference of the above model, we make use of the `BayesGP` package:

```{r}
mod <- BayesGP::model_fit(
    y ~ f(
      x,
      model = "sgp",
      region = c(0,10),
      freq = 2,
      k = 10, # number of knots
      sd.prior = list(param = list(u = 2, alpha = 0.1), h = 1)
    ) +
      f(index, model = "iid", sd.prior = 1),
    data = data,
    family = "Poisson"
  )
```


We can take a quick look at the posterior summary:

```{r}
summary(mod)
```

We can also obtain the posterior of $g$ at any location of interest:

```{r}
post_g <- predict(mod, newdata = data.frame(x = location_of_interest), variable = "x", include.intercept = FALSE)
head(post_g)
```


Take a look at the plot of them:

```{r}
plot(location_of_interest, true_f(location_of_interest),
     type = "l", col = "black",
     pch = 20, cex = 0.5,
     ylab = "y", xlab = "x")
lines(x = location_of_interest, y = (post_g$mean), col = "blue", lwd = 1, lty = 2)
polygon(c(location_of_interest, rev(location_of_interest)),
        c(post_g$q0.025, rev(post_g$q0.975)),
        col = adjustcolor("blue", alpha.f = 0.2), border = NA)
legend("topright", legend = c("True function", "Posterior mean"),
       col = c("black", "blue"), lty = c(1, 2), lwd = c(1, 1))
```



The quantiles reported from `predict` can be easily modified by setting the `quantiles` argument. For example, to obtain the $0.05$ and $0.95$ quantiles, we can set `quantiles = c(0.05, 0.95)`.

```{r}
post_g <- predict(mod, newdata = data.frame(x = location_of_interest), variable = "x", include.intercept = FALSE, quantiles = c(0.05, 0.95))
head(post_g)
```


We can also just obtain the raw samples of $g$ at these locations:
```{r}
post_g_raw <- predict(mod, newdata = data.frame(x = location_of_interest), variable = "x", only.samples = TRUE, include.intercept = FALSE)
plot(location_of_interest, true_f(location_of_interest),
     type = "l", col = "black",
     pch = 20, cex = 0.5,
     ylab = "y", xlab = "x")
matlines(location_of_interest, post_g_raw[,2:12], col = "pink", lty = 2, lwd = 0.5)
```













